Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-07-08 appendix.tex
edits to reply to referee's point 1

\subsection{Spurious growth of adiabatic perturbations when
  $\zeta=0$}\label{analytic_adia}  
A limitation of the reduced model described in Appendix
\ref{adia_improve}, \S\ref{sec:simplified} and used in \S\ref{analytical}, is that it
cannot be employed for adiabatic flow when there is vertical shear, even
if $h/\khat\ll1$.  We explain this by setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\zeta=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
{\bf shows that} $\hat{\upsilon}^2$ is complex for real $\khat$, which
indicates instability {\bf for any value of  $\gamma>1$. This is inconsistent
  with the second Solberg-Hoiland criterion  
(Eq. \ref{solberg2}), which implies that instability requires the disk
to be close to neutral stratification (i.e. $|\gamma-1|\ll 1$ for a
vertically isothermal disk).} 
%This is contradicts the second Solberg-Hoiland criterion 
%(Eq. \ref{solberg2}), which imply instability for adiabatic
%perturbations in a thin, vertically isothermal disk would require 
%$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because we have retained the global radial temperature 
gradient to obtain vertical shear, 
but have ignored it elsewhere in the linear problem (as well as the
background radial density gradient). Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

>>>>

\subsection{Spurious growth of adiabatic perturbations when
  $\zeta=0$}\label{analytic_adia}  
A limitation of the reduced model described in Appendix
\ref{adia_improve}, \S\ref{sec:simplified} and used in \S\ref{analytical}, is that it
cannot be employed for adiabatic flow when there is vertical shear, even
if $h/\khat\ll1$.  We explain this by setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\zeta=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
{\bf shows that} $\hat{\upsilon}^2$ is complex for real $\khat$, which
indicates instability {\bf for any value of  $\gamma>1$. This stability condition is inconsistent
  with the second Solberg-Hoiland criterion  
(Eq. \ref{solberg2}), which states that instability requires the disk
to be close to neutral stratification (i.e. $|\gamma-1|\ll 1$ for a
vertically isothermal disk).} 
%This is contradicts the second Solberg-Hoiland criterion 
%(Eq. \ref{solberg2}), which imply instability for adiabatic
%perturbations in a thin, vertically isothermal disk would require 
%$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This {\bf spurious growth in the $\zeta = 0$ model arises}
 because we have retained the global radial temperature 
gradient to obtain vertical shear, 
but have ignored it elsewhere in the linear problem (as well as the
background radial density gradient). Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$, {\bf not for adiabatic perturbations}.   %and small scale

### 2015-07-08 setup.tex
edits to reply to referee's point 1

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{subequations}\begin{align}
  \kappa^2 - \frac{1}{\rho C_P}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0 \label{second_SH} 
\end{align}\end{subequations}
\citep{tassoul78}. %, where $j=r^2\Omega$ is the specific angular
%momentum. %and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  
For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is easily satisfied. 
Thus, we consider the second
criterion. 
Without loss of generality (since Eq. \ref{second_SH} is even in $z$), we consider $z>0$
so that $\p_zP<0$. With $\p_r j^2\simeq r^3\OmK^2$ and
using Eq. \ref{vertical_shear} for $\p_zj^2$  we find that
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters, the right hand
side is $O(h^2) \lesssim 10^{-2}$. The left hand side is positive and order
unity ($O(0.1)$ at least) if the disk is convectively stable, and not
very close to convective instability.  This stable stratification is
expected for irradiated disks  \citep{chiang97}. 
Thus adiabatic disturbances are stable to a disk's vertical shear,
explaining why the VSI requires cooling.

>>>>

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{subequations}\begin{align}
  \kappa^2 - \frac{1}{\rho C_P}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0 \label{second_SH} 
\end{align}\end{subequations}
\citep{tassoul78}. %, where $j=r^2\Omega$ is the specific angular
%momentum. %and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  
For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is easily satisfied. 
Thus, we consider the second
criterion. 
Without loss of generality (since Eq. \ref{second_SH} is even in $z$), we consider $z>0$
so that $\p_zP<0$. With $\p_r j^2\simeq r^3\OmK^2$ and
using Eq. \ref{vertical_shear} for $\p_zj^2$  we find that
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters, the right hand
side is $O(h^2) \lesssim 10^{-2}$. The left hand side is positive and order
unity %if the disk is convectively stable, and not
%very close to convective instability  {\bf($\gamma-\Gamma\geq0.1$ in
  %the disk models considered)}. 
  {\bf in our disk models (with $\gamma-\Gamma\geq0.1$)
  implying strong stability to convection, see Eq.\ \ref{nzsq_def}.}
  This stable stratification is
expected for irradiated disks  \citep{chiang97}. 
Thus, adiabatic disturbances {\bf in a standard, irradiated disk}
 are stable to a disk's vertical shear,
%{\bf for typical values of  $\gamma-\Gamma$},  
explaining why the VSI requires cooling.
